---
title: "DESeq_Script"
author: "Chris"
date: "31/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




#Files required to run:


BIOL448J_Functions.R 
pairwise.adonis.r
my_colours.R 
Prenatal_alcohol_exposure_16S_otu_table.txt
Prenatal_alcohol_exposure_metadata.txt



### Details about the user input variables:
If you would like to group the data by a varible (can contain >2 options to filter by), please specifiy this in the "highlight_var" input. 

"Variable_to_filter_by" and "filter_by" are inputs requried if you would like to analyze a subset of your data. You can filter by up to 2 variables. If you do not require both or either filtering options, leave blank spaces. However, upon usage, "Variable_to_filter_by" requires the column name of the data intended to filter, and "filter_by" requires the string within that column to subset. 

Ex, highlight_var  = "prenatal_group"
    variable_to_filter_by1 = "sex"
    filter_by1 = "Female"
    variable_to_filter_by2 = "drug_group"
    filter_by2 = "No drug"
```{r, Input Variabels, include=TRUE, echo=FALSE}
highlight_var  = "prenatal_group"
variable_to_filter_by1 = ""
filter_by1 = ""
variable_to_filter_by2 = "drug_group"
filter_by2 = "No drug"

```


```{r, Set variables and please dont change, include=TRUE, echo=FALSE}

otu_file       = "Prenatal_alcohol_exposure_16S_otu_table.txt"
metadata_file  = "Prenatal_alcohol_exposure_metadata.txt"
files_are_csv  = FALSE
sample_ID_col  = "MI_ID"
otu_as_rows    = TRUE
otu_col_to_ignore = c(1:2,68:75)
otu_row_to_ignore = c()

```

Loading libraries and functions from other scripts
```{r, Source Functions, include=TRUE, echo=FALSE}


#if (!requireNamespace("BiocManager", quietly = TRUE))
#install.packages("BiocManager")

#BiocManager::install("DESeq2")
#BiocManager::install("phyloseq")

library(tidyverse)
library(vegan)
library(gridExtra)
library(labdsv)
library(indicspecies)




source("BIOL448J_Functions.R") #Loading functions written by Illan Rubin (2019 TA for BIO448). Used for alpha diversity.
source("pairwise.adonis.r") #Loading parwiseadonis function also provided by Illan Rubin.
source("my_colours.R")  #loading subsets of manual colours used for plotting.
```

Creates a unique dataset name based on the user filter variables specified above
```{r, Dataset Name, include=TRUE, echo=FALSE}
dataset_name = paste("PAE-",highlight_var,"-",filter_by1,filter_by2,sep = "")
```

Reading in input files (count matrix and metadata)
```{r, Reading Files Into R, include=TRUE, echo=FALSE}
  otu_table = read.table(otu_file,header = TRUE,sep = "\t")
  metadata_table = read.table(metadata_file,header = TRUE,sep = "\t")
  rownames(otu_table) <- paste("OTU",otu_table[,1],sep = "")
```


```{r, Fixing Data so Metadata Matches OTU, include=TRUE, echo=FALSE}

PAE_otu_t <- as.data.frame(t(otu_table[,(3:67)]))
PAE_taxa <- as.matrix((otu_table[,(68:75)]))
PAE_otu_t$MI_ID <- rownames(PAE_otu_t)

#sorting by MI_ID columns

PAE_metadata_sorted <- metadata_table[order(metadata_table$MI_ID),]
PAE_otu_t_sorted <- PAE_otu_t[order(PAE_otu_t$MI_ID),]


#removing the blank13 and 40 from OTUdata

PAE_metadata_sorted <- PAE_metadata_sorted[-(1),]
PAE_metadata_sorted <- PAE_metadata_sorted[-(40),]

#combined metadata and OTU data and specifying vectors for col and metadata used later on.
PAE <-as.data.frame(cbind(PAE_metadata_sorted,PAE_otu_t_sorted))
PAE <- PAE[,-ncol(PAE)]
PAE_meta_col <-c(1,2,3,4,5,6,7,8)
PAE_otu_col <-9:440 



```


The filtering step based on user filter variables at the beginning.
```{r, Filtering by user input, include=TRUE, echo=FALSE}

if (filter_by1 != "" && filter_by2 != ""){
  PAE = PAE %>%
  filter(PAE[[variable_to_filter_by1]] == filter_by1) 
  
  PAE = PAE %>%
  filter(PAE[[variable_to_filter_by2]] == filter_by2)
      
  
} else if (filter_by1 != ""){
      PAE = PAE %>%
  filter(PAE[[variable_to_filter_by1]] == filter_by1)
      
  
} else if (filter_by2 != ""){
      PAE = PAE %>%
  filter(PAE[[variable_to_filter_by2]] == filter_by2)
      
} 
```


```{r, Calculating Alpha Diveristy Measures, include=TRUE, echo=FALSE}
alphameasures = run_all_diversity(PAE,species_col = -PAE_meta_col, filter_by = highlight_var)
```

```{r, Plotting Alpha Diveristy Measures, include=TRUE, echo=FALSE}
my_colours2 = c("#98984d",
              "#b3669e")
richness_index = ggplot(alphameasures,aes(get(highlight_var), richness, fill = get(highlight_var)))+
  geom_bar(stat = "identity",position = "dodge",width = 0.7)+
  lims(y=c(0,500))+
  theme_classic()+
  theme(text = element_text(face = "bold",size=15))+
  ggtitle("Richness")+scale_fill_manual(values = my_colours2)

shannons_index = ggplot(alphameasures,aes(get(highlight_var), shannons, fill = get(highlight_var)))+
  geom_bar(stat = "identity",position = "dodge",width = 0.7)+
  lims(y=c(0,10))+
  theme_classic()+
  theme(text = element_text(face = "bold",size=15))+
  ggtitle("Shannons")+
  scale_fill_manual(values = my_colours2)


invsim_index = ggplot(alphameasures,aes(get(highlight_var), inv_simpsons, fill = get(highlight_var)))+
  geom_bar(stat = "identity",position = "dodge",width = 0.7)+
  lims(y=c(0,35))+
  theme_classic()+
  theme(text = element_text(face = "bold",size=15))+
  ggtitle("Inverse Simpsons")+
  scale_fill_manual(values = my_colours2)

chao1_index = ggplot(alphameasures,aes(get(highlight_var), chao1, fill = get(highlight_var)))+
  geom_bar(stat = "identity",position = "dodge",width = 0.7)+
  lims(y=c(0,500))+
  theme_classic()+
  theme(text = element_text(face = "bold",size=15))+
  ggtitle("Chao1")+
  scale_fill_manual(values = my_colours2)


pdf(paste(dataset_name,"-Alpha Diveristy.pdf"))
    
grid.arrange(richness_index,shannons_index,invsim_index,chao1_index,nrow = 2,ncol = 2)
 
dev.off()
```



```{r, Plotting Stress values across Dimensions, include=TRUE, echo=FALSE}
PAE_MDSdim<-checkMDSdim((PAE[-PAE_meta_col]),iter = 5)

plot_stress = ggplot(PAE_MDSdim, aes(dimension, stress))+
  geom_point(size = 2)+
  theme_classic()+
  theme(text = element_text(size=15,face = "bold"))
```

```{r, NMDS plot, include=TRUE, echo=FALSE }
PAE_MDS <- metaMDS(PAE[-PAE_meta_col])
PAE_MDSmeta = data.frame(cbind(PAE[,PAE_meta_col],PAE_MDS$points))


plot_MDS = ggplot(PAE_MDSmeta, aes(x = MDS1,y = MDS2))+
  geom_point(aes_string(color=highlight_var),size = 3)+
  coord_fixed()+
  theme_classic()+
  theme(text = element_text(size = 15, face = "bold"))




if (highlight_var != ""){
  MDS_centroid = calculate_ellipse(PAE_MDSmeta,filter_by = highlight_var)
  plot_MDS = plot_MDS   +   geom_path(data=MDS_centroid,mapping = aes(colour = filter),size =1.25)

}

```

```{r, Relative Abundance Analysis, include=TRUE, echo=FALSE}
meta <- PAE[,PAE_meta_col]
otu <- PAE[,-PAE_meta_col]

meta$total_abundance <- rowSums(PAE[,-PAE_meta_col])

otu_and_metadata <- as.data.frame(cbind(meta,otu))

PAE_grouped <- gather(otu_and_metadata, key = "ASV",value = "abundance",-(1:10))


PAE_grouped$legend = paste(PAE_grouped$prenatal_group, PAE_grouped$sex)

PAE_taxa <- as.data.frame(PAE_taxa)
PAE_taxa$ASV <- rownames(PAE_taxa)
PAE_sum <- inner_join(PAE_grouped,PAE_taxa,by = "ASV")
PAE_sum$relative_abundance <- PAE_sum$abundance/PAE_sum$total_abundance

```

```{r, Plotting Relative Abundance for top 50 grouped by phylum, include=TRUE, echo=FALSE}


PAE_taxa_abundance <- colSums(PAE[,-PAE_meta_col])

N=50

top_N_taxa <- names(sort(PAE_taxa_abundance,decreasing = TRUE)[1:N])

PAE_top_50 <- filter(PAE_sum, ASV %in% top_N_taxa)



Phylum_counts <- PAE_top_50 %>%
  group_by(Rank2,MI_ID,prenatal_group) %>%
  summarise(sum =sum(relative_abundance))

#names(Phylum_counts)[2] <- "Phylum" 

highlight_var = "prenatal_group"
plot1 = ggplot(Phylum_counts,aes(x=MI_ID,y=sum,fill=Rank2))+
  geom_bar(stat = "identity",color="black") +
  facet_grid(as.formula(paste("~",highlight_var)),scales = "free", space = "free")+
  theme(axis.text = element_text(face="bold",size= 7),
        text = element_text(face = "bold",size = 12),
        axis.text.x.bottom = element_blank())+
  labs(x="", y= " Relative Abundance (%)")+
  scale_fill_manual(values = c("#c9578c",
"#85a040",
"#8475ca",
"#cb7041",
"#4bae8d"))

ggsave("grouped_by_phylum.jpeg", plot = plot1)

```

```{r, Plotting Relative Abundance for the top 50 OTUs, include=TRUE, echo=FALSE}

PAE_taxa_abundance <- colSums(PAE[,-PAE_meta_col])

N=50

top_N_taxa <- names(sort(PAE_taxa_abundance,decreasing = TRUE)[1:N])

PAE_top_25 <- filter(PAE_sum, ASV %in% top_N_taxa)


Phylum_counts <- PAE_top_25 %>%
  group_by(legend) %>%
  summarise(sum =sum(abundance))

names(Phylum_counts)[2] <- "Phylum" 

#to group by sex too, use ~legend instead of fomula.as(paste())
ggplot(PAE_top_25,aes(x=MI_ID,y=relative_abundance,fill=Rank6))+
  geom_bar(stat = "identity",color="black") +
  facet_grid(~legend,scales = "free", space = "free")+
  theme_bw()+
  theme(axis.text = element_text(face="bold",size= 7),
        text = element_text(face = "bold",size = 12),
        axis.text.x.bottom = element_text(face = "bold",size = 11),
        legend.position = "none",
        axis.title.x=element_blank(),
        axis.ticks.x=element_blank())+
  labs(x="", y= " Relative Abundance (%)")+
  scale_fill_manual(values = c("#b78226",
"#5e67d9",
"#3cc755",
"#c66ae2",
"#72be3e",
"#8b45b5",
"#a5bb36",
"#d53ea4",
"#41932d",
"#df316d",
"#5ec16b",
"#a04395",
"#d8b42e",
"#5f87d8",
"#ea952a",
"#5c5b9f",
"#b8a73d",
"#b07fd9",
"#799837",
"#e678c5",
"#4cc892",
"#dd4c2b",
"#36bddc",
"#ac3d1b",
"#48c5bb",
"#cb3a48",
"#35997f",
"#b83c78",
"#529c5e",
"#e6648a",
"#357838",
"#b39cdc",
"#456d19",
"#a25f92",
"#71781e",
"#589ed3",
"#db742f",
"#277257",
"#e87159",
"#367042",
"#dd8aaf",
"#536c31",
"#a23e5a",
"#7bbd94",
"#8d495e",
"#a5bc72",
"#dd8081",
"#605f23",
"#d88860",
"#8b9151",
"#a0513d",
"#e1b05a",
"#a35a22",
"#dda875",
"#7e611b",
"#a07e47"))
```
