---
title: "DESeq_Script"
author: "Chris"
date: "31/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




#Files required to run:


BIOL448J_Functions.R 
pairwise.adonis.r
my_colours.R 
Prenatal_alcohol_exposure_16S_otu_table.txt
Prenatal_alcohol_exposure_metadata.txt



### Details about the user input variables:
If you would like to group the data by a varible (can contain >2 options to filter by), please specifiy this in the "highlight_var" input. 

"Variable_to_filter_by" and "filter_by" are inputs requried if you would like to analyze a subset of your data. You can filter by up to 2 variables. If you do not require both or either filtering options, leave blank spaces. However, upon usage, "Variable_to_filter_by" requires the column name of the data intended to filter, and "filter_by" requires the string within that column to subset. 

Ex, highlight_var  = "prenatal_group"
    variable_to_filter_by1 = "sex"
    filter_by1 = "Female"
    variable_to_filter_by2 = "drug_group"
    filter_by2 = "No drug"
```{r, Input Variabels, include=TRUE, echo=FALSE}
highlight_var  = "sex"
variable_to_filter_by1 = "prenatal_group"
filter_by1 = "Ethanol"
variable_to_filter_by2 = "drug_group"
filter_by2 = "No drug"

```


```{r, Set variables and please dont change, include=TRUE, echo=FALSE}

otu_file       = "Prenatal_alcohol_exposure_16S_otu_table.txt"
metadata_file  = "Prenatal_alcohol_exposure_metadata.txt"
files_are_csv  = FALSE
sample_ID_col  = "MI_ID"
otu_as_rows    = TRUE
otu_col_to_ignore = c(1:2,68:75)
otu_row_to_ignore = c()

```

Loading libraries and functions from other scripts
```{r, Source Functions, include=TRUE, echo=FALSE}


#if (!requireNamespace("BiocManager", quietly = TRUE))
#install.packages("BiocManager")

#BiocManager::install("DESeq2")
#BiocManager::install("phyloseq")

library(tidyverse)
library(vegan)
library(gridExtra)
library(phyloseq)
library(labdsv)
library(indicspecies)
library(DESeq2)



source("BIOL448J_Functions.R") #Loading functions written by Illan Rubin (2019 TA for BIO448). Used for alpha diversity.
source("pairwise.adonis.r") #Loading parwiseadonis function also provided by Illan Rubin.
source("my_colours.R")  #loading subsets of manual colours used for plotting.
```

Creates a unique dataset name based on the user filter variables specified above
```{r, Dataset Name, include=TRUE, echo=FALSE}
dataset_name = paste("PAE-",highlight_var,"-",filter_by1,filter_by2,sep = "")
```

Reading in input files (count matrix and metadata)
```{r, Reading Files Into R, include=TRUE, echo=FALSE}
  otu_table = read.table(otu_file,header = TRUE,sep = "\t")
  metadata_table = read.table(metadata_file,header = TRUE,sep = "\t")
  rownames(otu_table) <- paste("OTU",otu_table[,1],sep = "")
```


```{r, Fixing Data so Metadata Matches OTU, include=TRUE, echo=FALSE}

PAE_otu_t <- as.data.frame(t(otu_table[,(3:67)]))
PAE_taxa <- as.matrix((otu_table[,(68:75)]))
PAE_otu_t$MI_ID <- rownames(PAE_otu_t)

#sorting by MI_ID columns

PAE_metadata_sorted <- metadata_table[order(metadata_table$MI_ID),]
PAE_otu_t_sorted <- PAE_otu_t[order(PAE_otu_t$MI_ID),]


#removing the blank13 and 40 from OTUdata

PAE_metadata_sorted <- PAE_metadata_sorted[-(1),]
PAE_metadata_sorted <- PAE_metadata_sorted[-(40),]

#combined metadata and OTU data and specifying vectors for col and metadata used later on.
PAE <-as.data.frame(cbind(PAE_metadata_sorted,PAE_otu_t_sorted))
PAE <- PAE[,-ncol(PAE)]
PAE_meta_col <-c(1,2,3,4,5,6,7,8)
PAE_otu_col <-9:440 



```


The filtering step based on user filter variables at the beginning.
```{r, Filtering by user input, include=TRUE, echo=FALSE}
set.seed(192)
if (filter_by1 != "" && filter_by2 != ""){
  PAE = PAE %>%
  filter(PAE[[variable_to_filter_by1]] == filter_by1) 
  
  PAE = PAE %>%
  filter(PAE[[variable_to_filter_by2]] == filter_by2)
      
  
} else if (filter_by1 != ""){
      PAE = PAE %>%
  filter(PAE[[variable_to_filter_by1]] == filter_by1)
      
  
} else if (filter_by2 != ""){
      PAE = PAE %>%
  filter(PAE[[variable_to_filter_by2]] == filter_by2)
      
} 

#highlight_var = "legend"
#PAE_metadata_sorted$legend = paste(PAE_metadata_sorted$sex,PAE_metadata_sorted$prenatal_group)

#Running permanova 
PAE_permanova <- adonis(as.formula(paste("PAE[,-PAE_meta_col] ~",highlight_var)),data=PAE[,PAE_meta_col])
PAE_permanova

pairwise_table <- table(pairwise.adonis(PAE[,-PAE_meta_col],PAE[,PAE_meta_col][[highlight_var]]))
```

```{r, gathering all data in phyloseq format, include=TRUE, echo=FALSE}

otu_plus1 <- PAE[,-PAE_meta_col]+1   #Adding 1 to each value as a pseudocount 
PAE_otu_phyloseq <- otu_table(otu_plus1,taxa_are_rows = FALSE)
PAE_metadata_phyloseq <- sample_data(PAE[,PAE_meta_col])
PAE_taxonomy_phyloseq <- tax_table(PAE_taxa)

#Creating a phyloseq object for the DEseq analysis.
PAE_phyloseq <- phyloseq(PAE_otu_phyloseq,PAE_metadata_phyloseq,PAE_taxonomy_phyloseq)
sample_data(PAE_phyloseq)
```

```{r, Running DEseq Analysis , include=TRUE, echo=FALSE}

#Running DESEQ on the phyloseq object
PAE_DEseq <- phyloseq_to_deseq2(PAE_phyloseq, design = as.formula(paste("~",highlight_var)))
PAE_DEseq <- DESeq(PAE_DEseq,test="Wald",fitType="parametric")

PAE_DEseq_results <- results(PAE_DEseq)

#Combinding the results output with taxa information.
PAE_DEseq_results <- data.frame(cbind(PAE_DEseq_results,PAE_taxa))
```



```{r, Plotting differential abundance for all genus}
DEseqbeforefilter <- ggplot(PAE_DEseq_results,aes(x=Rank6,y=log2FoldChange,color=Rank6))+
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0)+
  theme(legend.position = "none")+
  theme(axis.text.x = element_text(angle = 90,hjust = 0,vjust = 0.5))

pdf(paste(dataset_name,"-DEseq before filtering.pdf"))

DEseqbeforefilter
 
dev.off()
```

Each point represents a specific OTU assignment. Multiple points are seen within a single genus (x axis) because that genus contains several unique species identifications each with a different Lag2FolgChange value. 
Filtering DESEQ results by significance of 0.05 and removing 
```{r, Filtering for differential abundances with pvalue < 0.05 and removing rows with missing taxomony calls, include=TRUE, echo=FALSE}
PAE_DEseq_results_sig <- filter(PAE_DEseq_results,padj<0.05)
PAE_DEseq_results_sig_narm <- filter(PAE_DEseq_results_sig)
```
only removed NAs to rank 6 because that is the genus level and we arent looking at the species level at all. mostly because they mostly identify as unculturem_bacterium
```{r, Taking the Mean of each genus, include=TRUE, echo=FALSE}
PAE_DEseq_results_genus_summary<-PAE_DEseq_results_sig_narm %>%
  group_by(Rank6,Rank5,Rank4,Rank3,Rank2) %>%
  select(Rank6,Rank5,Rank4,Rank3,Rank2,log2FoldChange,padj,baseMean,pvalue) %>%
  dplyr::summarise(mean_2_fold_change=mean(log2FoldChange),padj=combine_pvalues(padj),countmean = mean(baseMean),pvalue=combine_pvalues(pvalue))

```

```{r, Taking the Mean of each Family, include=TRUE, echo=FALSE}
PAE_DEseq_results_family_summary<-PAE_DEseq_results_sig_narm %>%
  group_by(Rank5,Rank4,Rank3,Rank2) %>%
  select(Rank5,Rank4,Rank3,Rank2,log2FoldChange,padj,baseMean,pvalue) %>%
  dplyr::summarise(mean_2_fold_change=mean(log2FoldChange),padj=combine_pvalues(padj),countmean = mean(baseMean),pvalue=combine_pvalues(pvalue))
```

```{r, Taking the Mean of each order, include=TRUE, echo=FALSE}
PAE_DEseq_results_order_summary<-PAE_DEseq_results_sig_narm %>%
  group_by(Rank4,Rank3,Rank2) %>%
  select(Rank4,Rank3,Rank2,log2FoldChange,padj,baseMean,pvalue) %>%
  dplyr::summarise(mean_2_fold_change=mean(log2FoldChange),padj=combine_pvalues(padj),countmean = mean(baseMean),pvalue=combine_pvalues(pvalue))
```

```{r, Taking the Mean of each class, include=TRUE, echo=FALSE}
PAE_DEseq_results_class_summary<-PAE_DEseq_results_sig_narm %>%
  group_by(Rank3,Rank2) %>%
  select(Rank3,Rank2,log2FoldChange,padj,baseMean,pvalue) %>%
  dplyr::summarise(mean_2_fold_change=mean(log2FoldChange),padj=combine_pvalues(padj),countmean = mean(baseMean),pvalue=combine_pvalues(pvalue))
```

```{r, Taking the Mean of each phyla, include=TRUE, echo=FALSE}
PAE_DEseq_results_phyla_summary<-PAE_DEseq_results_sig_narm %>%
  group_by(Rank2) %>%
  select(Rank2,log2FoldChange,padj,baseMean,pvalue) %>%
  dplyr::summarise(mean_2_fold_change=mean(log2FoldChange),padj=combine_pvalues(padj),countmean = mean(baseMean),pvalue=combine_pvalues(pvalue))
```


This is just some nonsense I saved for plotting difference comparisons due to different output tables.

PAE_DEseq_results_genus_summary$Rank6 <- gsub('GCA-900066575', 'NA', PAE_DEseq_results_genus_summary$Rank6)
PAE_DEseq_results_genus_summary$Rank6 <- gsub('uncultured_bacterium', 'NA ', PAE_DEseq_results_genus_summary$Rank6)
PAE_DEseq_results_genus_summary$Rank6 <- gsub('uncultured_Mollicutes_bacterium', 'NA  ', PAE_DEseq_results_genus_summary$Rank6)
PAE_DEseq_results_genus_summary[15,1] <- "NA    "
PAE_DEseq_results_genus_summary[16,1] <- "NA     "
PAE_DEseq_results_genus_summary[17,1] <- "NA      "
                        Male and female
                                c("#0dc34e",
                                "#C011D7",
                                "#e24223",
                                "#FFA52C",
                                "#000000")
                                
                                Males
                                c("#2F3CBE",
                                "#0dc34e",
                                "#e24223",
                                "#86007D")
                                
                                females
                                c(
                                "#0dc34e",
                                "#074E24",
                                "#e24223",
                                "#349CA5")

```{r, Plotting differential abundance at the genus level}
PAE_DEseq_results_genus_summary[10,1] <- "NA    "
PAE_DEseq_results_genus_summary[9,1] <- "NA     "
PAE_DEseq_results_genus_summary[8,1] <- "NA      "




DEseqafterfiltergenus <- ggplot(PAE_DEseq_results_genus_summary,aes(x=reorder(Rank6, -mean_2_fold_change),y=mean_2_fold_change,color = Rank2))+
  geom_point(size=3.5) +
  geom_hline(yintercept = 0,size=1)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = -90,hjust = 0,vjust = 0.5),
        text = element_text(face = "bold",size=15),
        axis.text.x.bottom  = element_text(size=12,angle = -45),)+
  scale_color_manual(values = c(
                                "#0dc34e",
                                "#074E24",
                                "#e24223",
                                "#349CA5"))+
  labs(x="Genus",y="Log2 Fold Change",color= "Phylum")+
  ggtitle(paste(filter_by1))

ggsave(paste(dataset_name,"-DEseq after filtering (genus).png"),plot= DEseqafterfiltergenus)

```

```{r, Plotting differential abundance at the family level}

PAE_DEseq_results_family_summary <- PAE_DEseq_results_family_summary[-c(11:12),]

DEseqafterfilterfamily <- ggplot(PAE_DEseq_results_family_summary,aes(x=reorder(Rank5, -mean_2_fold_change),y=mean_2_fold_change,color = Rank2))+
  geom_point(size=3.5) +
  geom_hline(yintercept = 0,size=1)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = -90,hjust = 0,vjust = 0.5),
        text = element_text(face = "bold",size=12),
        axis.text.x.bottom  = element_text(size=9,angle = -45))+
  scale_fill_manual(values = colours_12)+
  labs(x="Family",y="Average Log2 Fold Change",color= "Phylum")

pdf(paste(dataset_name,"-DEseq after filtering (family).pdf"))

DEseqafterfilterfamily
 
dev.off()
```

```{r, plotting differential abundance at the phylum level}
DEseqafterfilterphylum <- ggplot(PAE_DEseq_results_phyla_summary,aes(x=reorder(Rank2, -mean_2_fold_change),y=mean_2_fold_change,color = Rank2))+
  geom_point(size=3.5) +
  geom_hline(yintercept = 0,size=1)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = -90,hjust = 0,vjust = 0.5),
        text = element_text(face = "bold",size=12),
        legend.position = "none",
        axis.text.x.bottom  = element_text(size=9,angle = -45))+
  scale_color_manual(values = c("#426df7", 
                                "#0dc34e",
                                "#e24223",
                                "#86007D"))+
  labs(x="Phylum",y="Log2 Fold Change",color= "Phylum")+
  ggtitle(paste(filter_by1))

ggsave(paste(dataset_name,"-DEseq after filtering (phylum).jpeg"),plot= DEseqafterfilterphylum)

```